---
title: 'DESeq HSC with "Activation" genes normalization'
output:
  html_document:
    fig_height: 6
    fig_width: 10
  pdf_document: default
---

Description

# Date of anlaysis

```{r timestamp}
date()
```

---

# Preparing input files

- Loading library

```{r loading_lib, message=FALSE}
library("BiocParallel")
register(MulticoreParam(10))
library('DESeq2')
```

- Set Filename and directory

```{r filename}
csv_path='SampleSheet_HSC_stranded_Proliferation.tsv'
ref_condition="A"
outdir="Proliferation"
### Example sample sheet
# Sample sheet is SPACE-DELIMITED, with header
# Example
# sampleName fileName condition1
# HET_721 HET_721.HQ.counts HET
# HET_729 HET_729.HQ.counts HET

sampleTable <- read.table(csv_path,header=TRUE)
sample_dir='../mapping/'
```

* Read files and generate "dds"

```{r deseq, message=FALSE}
## DESIGN
# Single factor -> CONDITION
#ddsHTSeq <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,directory = sample_dir, design= ~ condition)
# MULTI FACTOR -> Condition + proliferation status
ddsHTSeq <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,directory = sample_dir, design= ~ condition + type)
##

# DESeq worker
dds <- DESeq(ddsHTSeq, parallel=TRUE)
# Plot dispersion
plotDispEsts(dds)

## Pre-filtering
# All samples > 1 in gene count
dds<-dds[rowSums(counts(dds)) > 1,] 
# Extracting counts
raw.counts <- as.data.frame(counts( dds ))
normalized.counts <- as.data.frame(counts( dds, normalized=TRUE ))

# Relevel the condition
dds$type <- relevel(dds$type, ref=ref_condition)
res <- results(dds, parallel=TRUE)
head(res)
# Summary of DEGs
summary(res)

# Stabilizing the variance (takes long...)
rld <- rlog(dds)
```

- Gene annotation Mouse

```{r annotation, message=F}
#retrieve gene data
library("biomaRt")
gene_set = rownames(res)
ensembl=useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                dataset = "mmusculus_gene_ensembl",
                host = "feb2014.archive.ensembl.org",
                path = "/biomart/martservice")
genes_information = getBM(c("mgi_symbol",
                            "entrezgene",
                            "uniprot_genename",
                            "description",
                            "gene_biotype",
                            "chromosome_name",
                            "strand",
                            "start_position",
                            "end_position",
                            "ensembl_gene_id"),
                          "ensembl_gene_id",
                          gene_set, ensembl)
```

- Generate files for GSEA

```{r export_for_GSEA}
gene_set = rownames(normalized.counts)
res_information = getBM(c("ensembl_gene_id",
                          "mgi_symbol"),
                          "ensembl_gene_id",
                          gene_set, ensembl)
tmpfile = cbind(ensembl_gene_id = rownames(normalized.counts), normalized.counts)
rownames(tmpfile) <- NULL
gseafilename = merge(res_information, tmpfile, by = "ensembl_gene_id")[,-1]
gseafilename = gseafilename[gseafilename$mgi_symbol != "",]

write.table(gseafilename, paste(outdir, "HSC_Results_GSEA_input.tsv", sep = "/"), sep="\t", quote=F, row.names=F)

## Pre-Rank file based on log2FC
res_df = as.data.frame(res)
res_df = cbind(ensembl_gene_id = rownames(res_df), res_df)
rownames(res_df) <- NULL
rankGSEA = merge(res_information, res_df, by = "ensembl_gene_id")[,-1]
rankGSEA = rankGSEA[rankGSEA$mgi_symbol != "",]
rankGSEA = rankGSEA[order(rankGSEA$log2FoldChange, decreasing = F),]

write.table(rankGSEA[,c("mgi_symbol","log2FoldChange")], paste(outdir, "HSC_Results_GSEA_input.rnk", sep = "/"), sep="\t", quote=F, row.names=F)
```

- Generate "means" DF according to treatment (type A/B proliferation)

```{r means_df, message=F}
# Variable for the condition
populations = levels(dds$type)

#Generate 'means' dataframe
len = length(rowMeans(counts(dds,normalized=TRUE)[,dds$type == populations[19]]))
means = data.frame(matrix(NA, nrow = len))
count = 0
for (pop in populations){
  print(pop)
  tryCatch({means[pop] = rowMeans(counts(dds,normalized=TRUE)[,dds$type == pop])},
           error = function(e) {
             means[pop] = counts(dds,normalized=TRUE)[,dds$type == pop]
           })    
}
# Remove NA initial column from 'means'
means = means[,2:length(means)]
rownames(means) = rownames(normalized.counts)

means$ensembl_gene_id = rownames(means)
genes_with_information = merge(x = means, y = genes_information, by = "ensembl_gene_id", all.x = TRUE)
```

- Select significant genes

```{r sigGenes}
alpha=0.05
resSig <- res[which(res$padj <=alpha),]
resSig$ensembl_gene_id=rownames(as.data.frame(resSig))
downDEGs <- res[which(res$padj <=alpha & res$log2FoldChange < 0),]
upDEGs <- res[which(res$padj <=alpha & res$log2FoldChange > 0),]
```

- Write DEGs results with annotation

```{r DEGs_results}
# Significant gene annotation
resSig_annot = merge(as.data.frame(resSig), genes_with_information, by="ensembl_gene_id")
write.table(resSig_annot, paste(outdir, "HSC_DEGs_annotation.xls", sep = "/"), sep="\t", quote=F, row.names=F)

# UP genes
write.table(resSig_annot[which(resSig_annot$log2FoldChange>=0),], paste(outdir,"HSC_DEGs_UP_annotation.xls", sep = "/"), sep="\t", quote=F, row.names=F)
# DOWN genes
write.table(resSig_annot[which(resSig_annot$log2FoldChange<=0),], paste(outdir, "HSC_DEGs_DOWN_annotation.xls", sep = "/"), sep="\t", quote=F, row.names=F)
```

- Write ALL genes with DESeq statistics (Useful for GSEA)

```{r ALL_results}
# All genes with annotation
gene_set = rownames(res)
res_information = getBM(c("ensembl_gene_id",
                          "mgi_symbol"),
                          "ensembl_gene_id",
                          gene_set, ensembl)
res2write = cbind(ensembl_gene_id = rownames(res), as.data.frame(res))
res2write = merge(res2write, res_information, by = "ensembl_gene_id")
write.table(res2write, paste(outdir,"HSC_DESeq_Results.xls", sep="/"), sep="\t", quote=F, row.names=F)
```

- Preparing data for plot

```{r preparing_data_for_plot, echo=F}

# Which genes are significant
ens_genesSig=rownames(resSig)

# Extract annotation
geneSig = genes_with_information[genes_with_information$ensembl_gene_id %in% ens_genesSig,]

# Create DF for means with geneSig annotation
means$significant[means$ensembl_gene_id %in% ens_genesSig] <- 1
means$significant[!(means$ensembl_gene_id %in% ens_genesSig)] <- 0
genes_with_information$significant[genes_with_information$ensembl_gene_id %in% ens_genesSig] <- 1
genes_with_information$significant[!(genes_with_information$ensembl_gene_id %in% ens_genesSig)] <- 0
```

# Visualization

- Plot counts of specific genes
- "TNF","IL1B","IL6","CXCL1","CXCL10","INSR","LOXL2","CCL2","COL1A1","ACTA2"
## IFNG - ENSMUSG00000055170
-> 0 counts in all the samples

```{r plotcounts_specific, message = F}
# Plot contus of specific genes
library(ggplot2)
### List of specific genes
## From Marica presentation
gene_list = c("TNF","IL1B","IL6","CXCL1","CXCL10","INSR","LOXL2","CCL2","COL1A1","ACTA2","PECAM1")
## some cycline
#gene_list = c("CCNB1","CCNB2")

ensgene_list = sapply(gene_list, function(x){
  unique(genes_with_information[which(toupper(genes_with_information$mgi_symbol)==x),]$ensembl_gene_id)
  })
# Plot gene counts
#par(mfrow=c(3,1),mar=c(3,5,3,1))
plots = lapply(names(ensgene_list),function(x){
  if(length(ensgene_list[[x]]) != 0){
    data <- plotCounts(dds, gene=ensgene_list[[x]], intgroup=c("condition","type"), normalized=T, returnData=TRUE)
    ggplot(data, aes(x=type, y=count, colour = condition)) +
      #scale_y_log10() + 
      #scale_colour_continuous(high= "red", low =  "black") +
      geom_point(position=position_jitter(width=.1,height=0), size=5)+
      geom_text(aes(label=rownames(data)),
                #position = position_jitter(width=.5,height=0)+
                hjust=2, 
                vjust=0)+
      ggtitle(paste(as.character(x),as.character(ensgene_list[[x]]), sep = " - "))
    }
})
data <- plotCounts(dds, gene=ensgene_list[[1]], intgroup=c("type","condition"), normalized=T, returnData=TRUE)
plots
# Gene list are significant?
gene_list_res=as.data.frame(res[rownames(res) %in% unique(genes_with_information[which(toupper(genes_with_information$mgi_symbol) %in% gene_list ),]$ensembl_gene_id),])
gene_list_res=cbind(ensembl_gene_id=rownames(gene_list_res), gene_list_res)
gene_list_res = merge(x = res_information, y = gene_list_res, by = "ensembl_gene_id")
gene_list_res$significant=gene_list_res$padj<alpha
gene_list_res
```

- HSC markers plot

```{r plotcounts_HSC_marker, message = F}
# Plot contus of specific genes
library(ggplot2)
### List of specific genes
### http://www.nature.com/nprot/journal/v10/n2/pdf/nprot.2015.017.pdf
### HIGH-retinoid --> GOOD HSC purification --> UP regulation of markers
### LOW-retinoid --> BAD HSC purification --> DOWN regulation of markers
### Activation marker genes
gene_list = c("LRAT","LHX2","HAND2","PDGFRB","VIM", "GFAP", "DES", "PPARG","TGFBR1","ACTA2","BAMBI")

ensgene_list = sapply(gene_list, function(x){
  unique(genes_with_information[which(toupper(genes_with_information$mgi_symbol)==x),]$ensembl_gene_id)
  })
# Plot gene counts
#par(mfrow=c(3,1),mar=c(3,5,3,1))
plots = lapply(names(ensgene_list),function(x){
  if(length(ensgene_list[[x]]) != 0){
    data <- plotCounts(dds, gene=ensgene_list[[x]], intgroup=c("condition","type"), normalized=T, returnData=TRUE)
    ggplot(data, aes(x=type, y=count, colour = condition)) +
      #scale_y_log10() + 
      #scale_colour_continuous(high= "red", low =  "black") +
      geom_point(position=position_jitter(width=.1,height=0), size=5)+
      geom_text(aes(label=rownames(data)),
                #position = position_jitter(width=.5,height=0)+
                hjust=2, 
                vjust=0)+
      ggtitle(paste(as.character(x),as.character(ensgene_list[[x]]), sep = " - "))
    }
})
data <- plotCounts(dds, gene=ensgene_list[[1]], intgroup=c("type","condition"), normalized=T, returnData=TRUE)
plots
# Gene list are significant?
gene_list_res=as.data.frame(res[rownames(res) %in% unique(genes_with_information[which(toupper(genes_with_information$mgi_symbol) %in% gene_list ),]$ensembl_gene_id),])
gene_list_res=cbind(ensembl_gene_id=rownames(gene_list_res), gene_list_res)
gene_list_res = merge(x = res_information, y = gene_list_res, by = "ensembl_gene_id")
gene_list_res$significant=gene_list_res$padj<alpha
gene_list_res

```

- PCA

```{r pca, message = F}
data_PCA <- plotPCA(rld, intgroup = c("type","condition"), returnData=TRUE)
percentVar <- round(100 * attr(data_PCA, "percentVar"))
library("ggplot2")
qplot(PC1, PC2, color=condition, data=data_PCA, label=rownames(data_PCA)) +
  #scale_colour_gradient(low="black", high="red") +
  geom_point(aes (shape = factor(type) ), size=4) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + geom_text(vjust = 1.7, nudge_x = 0.05, check_overlap = T)

```

- Scatter plot

```{r scatter, message = F}
library(ggplot2)
colScale <- scale_colour_manual(name = "significant", values = c("gray","red"))
ggplot(genes_with_information, aes(A, B, colour=as.factor(significant), alpha = as.factor(significant))) +
  geom_point() +
  colScale +
  # label
  #geom_text(aes(label=ifelse((significant==1),toupper(as.character(mgi_symbol)),'')),
  #          hjust=0, 
  #          vjust=0, 
  #          check_overlap = T) +
  scale_y_log10() +
  scale_x_log10()

```

- Gene clustering

```{r gene_clust, message = F}
# DEGs Gene
library("genefilter")
library("pheatmap")

mat <- assay(rld)[ which(rownames(rld) %in% rownames(resSig)),]
mat <- mat - rowMeans(mat)
# Create Gene name dataframe
#df_gene <- as.data.frame(rownames(mat))
#colnames(df_gene) <- "ensembl_gene_id"
#df_gene = merge(x = df_gene, y = genes_information, by = "ensembl_gene_id", all.x = TRUE)
# fix NA of BLANK values with ENSID
#idx_to_fix = which(df_gene$mgi_symbol == "" | is.na(df_gene$mgi_symbol))
#df_gene[idx_to_fix,]$mgi_symbol = as.vector(df_gene[idx_to_fix,]$ensembl_gene_id)
#rownames(mat) = unique(df_gene[match(df_gene$ensembl_gene_id,rownames(mat)),]$mgi_symbol)
# Create condition dataframe
df <- as.data.frame(colData(rld)[,"type"])
colnames(df) <- "type"
rownames(df) <- rownames(colData(rld))
# Plot the heatmap
pheatmap(mat, annotation_col = df, show_rownames = F)
```

