---
title: "GSEA Analysis on HSC DEGs"
author: "apietrelli"
date: "27/4/2016"
output: 
  html_document:
    fig_height: 8
    fig_width: 15
    theme: united
    toc: yes
---

```{r date}
date()
```

```{r library_loading, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(reshape2)
library(knitr)
outdir="../GSEA/GSEA_parsing/"
```

# GSEA Parsing

Find how many DEGs are in GSEA pathway enriched

```{r read_files, echo=FALSE}
# DEG file
DEG_file="RUV_Normalization_27042016/HSC_DEGs_annotation.xls"
DEG = read.csv(DEG_file, sep = "\t", header = T, stringsAsFactors = F)
# toupper mgi_symbol
DEG$mgi_symbol=toupper(DEG$mgi_symbol)
# remove NA in mgi_symbol
DEG = DEG[!is.na(DEG$mgi_symbol),]

# All Genes
AllGenes_file="RUV_Normalization_27042016/HSC_DESeq_Results.xls"
AllGenes = read.csv(AllGenes_file, sep = "\t", header = T, stringsAsFactors = F)
# toupper mgi_symbol
AllGenes$mgi_symbol=toupper(AllGenes$mgi_symbol)
# remove NA in mgi_symbol
AllGenes = AllGenes[!is.na(DEG$mgi_symbol),]
```

```{r GSEA_parsing}
GSEA_parse=function(GSEA_file,DEG,summary=0){
  
  GSEA=read.csv(GSEA_file, sep = "\t", header = T, stringsAsFactors = F)

  # Extract gene belonging to CORE ENRICHMENT
  GSEA = GSEA[GSEA$CORE.ENRICHMENT=="Yes",]
  GSEA$mgi_symbol = as.character(GSEA$GENE.SYMBOL)
  n_GSEA = nrow(GSEA)
  GeneSet = sub("^([^.]*).*", "\\1", basename(GSEA_file))
  # Join with DEGs
  res_data = merge(GSEA,DEG,"mgi_symbol")
  n_res_data = nrow(res_data)
  res_data[,"NAME"] = NULL 
  #print(res_data)
  #print(GSEA)
  #print(nrow(res_data))
  
    res_vec = data.frame(GeneSet=as.character(GeneSet),
                         n_GSEA=n_GSEA,
                         n_gene=n_res_data,
                         fraction=as.numeric((n_res_data/n_GSEA)*100))
  if(summary==0){
    if (nrow(res_data) > 0) {
      return(as.data.frame(c(res_vec,res_data)))
    }
  }
  else{
    return(res_vec)
  }
}

```

## Hallmark GSEA data parsing

```{r Hallmark_signal2noise}
## Hallmark
GSEA_dir="../GSEA/my_analysis.Gsea.1461755156354.Hallmark/"

# Store the geneset significance
HET_summary_GSEA_file="../GSEA/my_analysis.Gsea.1461755156354.Hallmark/gsea_report_for_HET_1461755156354.xls"
WT_summary_GSEA_file="../GSEA/my_analysis.Gsea.1461755156354.Hallmark/gsea_report_for_WT_1461755156354.xls"
HET_summary_GSEA = read.csv(HET_summary_GSEA_file, sep = "\t", header = T, stringsAsFactors = F)
WT_summary_GSEA = read.csv(WT_summary_GSEA_file, sep = "\t", header = T, stringsAsFactors = F)

HET_summary_GSEA = HET_summary_GSEA[,c("NAME","FDR.q.val")]
WT_summary_GSEA = WT_summary_GSEA[,c("NAME","FDR.q.val")]

colnames(HET_summary_GSEA) = c("GeneSet","FDR.q.val")
colnames(WT_summary_GSEA) = c("GeneSet","FDR.q.val")

DEG_HET = DEG[DEG$log2FoldChange>0,]
DEG_WT = DEG[DEG$log2FoldChange<0,]

# WT
output_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^HALL.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_WT)
  output_parsing = rbind(output_parsing,res_data)
  output_parsing = unique(output_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
WT_GSEA_parsing.Hallmark = merge(WT_summary_GSEA,output_parsing,"GeneSet")

# HET
output_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^HALL.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_HET)
  output_parsing = rbind(output_parsing,res_data)
  output_parsing = unique(output_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
HET_GSEA_parsing.Hallmark = merge(HET_summary_GSEA,output_parsing,"GeneSet")

# Summary
# WT
summary_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^HALL.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_WT,1)
  summary_parsing = rbind(summary_parsing,res_data)
  summary_parsing = unique(summary_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
WT_GSEA_summary.Hallmark = merge(WT_summary_GSEA,summary_parsing,"GeneSet")
# HET
summary_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^HALL.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_HET,1)
  summary_parsing = rbind(summary_parsing,res_data)
  summary_parsing = unique(summary_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
HET_GSEA_summary.Hallmark = merge(HET_summary_GSEA,summary_parsing,"GeneSet")

```

## Hallmark Results

### Hallmark Enriched in WT condition
```{r write_Hallmark_WT}
# WT 
kable(unique(WT_GSEA_summary.Hallmark[order(WT_GSEA_summary.Hallmark$FDR.q.val),1:5]))
```

### Hallmark Enriched in HET condition
```{r write_Hallmark_HET}
# HET
kable(unique(HET_GSEA_summary.Hallmark[order(HET_GSEA_summary.Hallmark$FDR.q.val),1:5]))
```

```{r write_Hallmark_files}
write.table(WT_GSEA_parsing.Hallmark, paste(outdir, "HSC_GSEA.WT.Hallmark.tsv", sep = "/"), sep="\t", quote=F, row.names=F)
write.table(HET_GSEA_parsing.Hallmark, paste(outdir, "HSC_GSEA.HET.Hallmark.tsv", sep = "/"), sep="\t", quote=F, row.names=F)
```

## KEGG GSEA data parsing

```{r KEGG_signal2noise}
GSEA_dir="../GSEA/KEGG.Gsea.1461765083398/"
output_parsing=NULL

# Store the geneset significance
HET_summary_GSEA_file=paste(GSEA_dir,"gsea_report_for_HET_1461765083398.xls", sep="")
WT_summary_GSEA_file=paste(GSEA_dir,"gsea_report_for_WT_1461765083398.xls", sep="")
HET_summary_GSEA = read.csv(HET_summary_GSEA_file, sep = "\t", header = T, stringsAsFactors = F)
WT_summary_GSEA = read.csv(WT_summary_GSEA_file, sep = "\t", header = T, stringsAsFactors = F)

HET_summary_GSEA = HET_summary_GSEA[,c("NAME","FDR.q.val")]
WT_summary_GSEA = WT_summary_GSEA[,c("NAME","FDR.q.val")]

colnames(HET_summary_GSEA) = c("GeneSet","FDR.q.val")
colnames(WT_summary_GSEA) = c("GeneSet","FDR.q.val")

DEG_HET = DEG[DEG$log2FoldChange>0,]
DEG_WT = DEG[DEG$log2FoldChange<0,]

# WT
output_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^KEGG.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_WT)
  output_parsing = rbind(output_parsing,res_data)
  output_parsing = unique(output_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
WT_GSEA_parsing.KEGG = merge(WT_summary_GSEA,output_parsing,"GeneSet")

# HET
output_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^KEGG.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_HET)
  output_parsing = rbind(output_parsing,res_data)
  output_parsing = unique(output_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
HET_GSEA_parsing.KEGG = merge(HET_summary_GSEA,output_parsing,"GeneSet")

# Summary
# WT
summary_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^KEGG.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_WT,1)
  summary_parsing = rbind(summary_parsing,res_data)
  summary_parsing = unique(summary_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
WT_GSEA_summary.KEGG = merge(WT_summary_GSEA,summary_parsing,"GeneSet")
# HET
summary_parsing=NULL
for (file in list.files(GSEA_dir,pattern = "^KEGG.*xls")){
  res_data=GSEA_parse(paste(GSEA_dir,file,sep=""),DEG_HET,1)
  summary_parsing = rbind(summary_parsing,res_data)
  summary_parsing = unique(summary_parsing)
}
output_parsing$GeneSet = as.character(output_parsing$GeneSet)
HET_GSEA_summary.KEGG = merge(HET_summary_GSEA,summary_parsing,"GeneSet")
```

## KEGG results

### KEGG Enriched in WT condition
```{r write_KEGG_WT}
# WT 
kable(unique(WT_GSEA_summary.KEGG[order(WT_GSEA_summary.KEGG$FDR.q.val),1:5]) )
```

### KEGG Enriched in HET condition

```{r write_KEGG_HET}
# HET
kable(unique(HET_GSEA_summary.KEGG[order(HET_GSEA_summary.KEGG$FDR.q.val),1:5]) )
```

```{r write_KEGG_files}
write.table(WT_GSEA_parsing.KEGG, paste(outdir, "HSC_GSEA.WT.KEGG.tsv", sep = "/"), sep="\t", quote=F, row.names=F)
write.table(HET_GSEA_parsing.KEGG, paste(outdir, "HSC_GSEA.HET.KEGG.tsv", sep = "/"), sep="\t", quote=F, row.names=F)
```


# Visualization

Prepare the input

```{r viz_input_preparation}
# Hallmark
WT_hist.Hallmark = melt(unique(WT_GSEA_parsing.Hallmark[,c("GeneSet","n_GSEA","n_gene","FDR.q.val")]),
                        id.vars = c("GeneSet","FDR.q.val"))
WT_hist.Hallmark$sample = rep("WT",nrow(WT_hist.Hallmark))
WT_hist.Hallmark$category = rep("Hallmark",nrow(WT_hist.Hallmark))

HET_hist.Hallmark = melt(unique(HET_GSEA_parsing.Hallmark[,c("GeneSet","n_GSEA","n_gene","FDR.q.val")]),
                         id.vars = c("GeneSet","FDR.q.val"))
HET_hist.Hallmark$sample = rep("HET",nrow(HET_hist.Hallmark))
HET_hist.Hallmark$category = rep("Hallmark",nrow(HET_hist.Hallmark))

# KEGG
WT_hist.KEGG = melt(unique(WT_GSEA_parsing.KEGG[,c("GeneSet","n_GSEA","n_gene","FDR.q.val")]),
                         id.vars = c("GeneSet","FDR.q.val"))
WT_hist.KEGG$sample = rep("WT",nrow(WT_hist.KEGG))
WT_hist.KEGG$category = rep("KEGG",nrow(WT_hist.KEGG))

HET_hist.KEGG = melt(unique(HET_GSEA_parsing.KEGG[,c("GeneSet","n_GSEA","n_gene","FDR.q.val")]),
                         id.vars = c("GeneSet","FDR.q.val"))
HET_hist.KEGG$sample = rep("HET",nrow(HET_hist.KEGG))
HET_hist.KEGG$category = rep("KEGG",nrow(HET_hist.KEGG))

# Merge to unique dataframe
data_GSEA = rbind(WT_hist.Hallmark, HET_hist.Hallmark,
                  WT_hist.KEGG, HET_hist.KEGG)

```

## Histogram plot for the enriched Gene Set and DEGs

Representation of how many DEGs are within the enriched Gene Set

```{r viz_plot_GSEA, fig.width=15, fig.height=10,echo=FALSE}
fdr=0.1
ggplot(data = data_GSEA[data_GSEA$FDR.q.val<=fdr,], aes(x=GeneSet, y=value, fill = variable)) +
  theme(axis.text.x=element_text(angle=30, hjust=1)) +
  geom_bar(stat = "identity", position = "dodge") + 
  geom_text(aes(label =value), 
            position=position_dodge(width=0.9),
            vjust=-0.25,
            size=3) +
  ggtitle(paste("GSEA enriched pathways and DEGs overlap (min 1 gene overlap)",fdr,sep=" - FDR<=")) +
  theme(plot.margin=unit(c(5,0,0,32),"mm")) +
  facet_grid(sample ~ category, scales = "free")

```

## Clustering for interesting pathways

```{r gene_clust, ignore = TRUE, include = FALSE, message = F}
# DEGs Gene
library("genefilter")
library("pheatmap")
library("DESeq2")
alpha=0.1

# Read rld as RDS object from DESeq2_RUV_2Step_norm
rld=readRDS("RUV_Normalization_27042016/HSC_RUV_rld.rds")

gene_signature = as.character(WT_GSEA_parsing.Hallmark$ensembl_gene_id[WT_GSEA_parsing.Hallmark$GeneSet=="HALLMARK_ALLOGRAFT_REJECTION"])

mat <- assay(rld)[ which(rownames(rld) %in% gene_signature),]
mat <- mat - rowMeans(mat)
# Create Gene name dataframe
#df_gene <- as.data.frame(rownames(mat))
#colnames(df_gene) <- "ensembl_gene_id"
#df_gene = merge(x = df_gene, y = genes_information, by = "ensembl_gene_id", all.x = TRUE)
# fix NA of BLANK values with ENSID
#idx_to_fix = which(df_gene$mgi_symbol == "" | is.na(df_gene$mgi_symbol))
#df_gene[idx_to_fix,]$mgi_symbol = as.vector(df_gene[idx_to_fix,]$ensembl_gene_id)
#rownames(mat) = unique(df_gene[match(df_gene$ensembl_gene_id,rownames(mat)),]$mgi_symbol)
# Create condition dataframe
df <- as.data.frame(colData(rld)[,"condition"])
colnames(df) <- "condition"
rownames(df) <- rownames(colData(rld))
# Plot the heatmap
pheatmap(mat, annotation_col = df, show_rownames = T, cluster_rows=T, cluster_cols=FALSE)
```

```{r extract_gene_clust, message = F}
library(gplots)

# Read rld as RDS object from DESeq2_RUV_2Step_norm
rld=readRDS("RUV_Normalization_27042016/HSC_RUV_rld.rds")

heatmap_patway=function(GSEA_parsing_data,GeneSetName,rld,DEG=1,keySize=1){
  # Extract DEG gene from geneset
  if (DEG==1){
    gene_signature = as.character(GSEA_parsing_data$ensembl_gene_id[GSEA_parsing_data$GeneSet==GeneSetName])
  }
  # All Genes
  else{
    gene_signature = as.character(GSEA_parsing_data$ensembl_gene_id)
  }
  mat <- assay(rld)[ which(rownames(rld) %in% gene_signature),]
  mat <- mat - rowMeans(mat)
  
  # choose a color palette for the heat map
  myheatcol <- rev(redgreen(75))
  
  # Change rownames
  GeneSymbol=NULL
  for (gene in rownames(mat)){
    g = as.character(unique(GSEA_parsing_data$GENE.SYMBOL[GSEA_parsing_data$ensembl_gene_id==gene]))
    GeneSymbol=c(GeneSymbol,g)
  }
  
  # draw the heat map with heatmap.2
  par(cex.main=1.2)
  heatmap.2(mat, main=GeneSetName,
            Colv = NA,
            dendrogram = "row",
            scale="row",
            col=myheatcol,
            density.info="none",
            trace="none",
            margins = c(7,8),
            labRow = GeneSymbol,
            keysize = keySize)
}

# HALLMARK GeneSet 
# Down regulated genes (WT)
DOWN.Hallmark_GeneSet = c("HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION",
                          "HALLMARK_INFLAMMATORY_RESPONSE",
                          "HALLMARK_KRAS_SIGNALING_UP",
                          "HALLMARK_COAGULATION")
for (i in DOWN.Hallmark_GeneSet){
  heatmap_patway(GSEA_parsing_data = WT_GSEA_parsing.Hallmark,
                 GeneSetName = i,
                 rld)
}


# KEGG GeneSet 
# Down regulated genes (WT)
DOWN.KEGG_GeneSet = c("KEGG_CELL_ADHESION_MOLECULES_CAMS",
                          "KEGG_ECM_RECEPTOR_INTERACTION",
                          "KEGG_ADHERENS_JUNCTION")
for (i in DOWN.KEGG_GeneSet){
  heatmap_patway(GSEA_parsing_data = WT_GSEA_parsing.KEGG,
                 GeneSetName = i,
                 rld)
}

```

# Specific pathway with no DEG but significantly enriched

## Pathway enriched in HET condition with no DEGs in overlap

```{r GeneSet_with_no_DEG, fig_height = 15, fig_width = 15}
# Read rld as RDS object from DESeq2_RUV_2Step_norm
rld=readRDS("RUV_Normalization_27042016/HSC_RUV_rld.rds")

# HALLMARK_ADIPOGENESIS
GeneSetName="HALLMARK_ADIPOGENESIS"
GSEA_file="../GSEA/my_analysis.Gsea.1461755156354.Hallmark/HALLMARK_ADIPOGENESIS.xls"
res_path = GSEA_parse(GSEA_file,AllGenes)

# Heatmap
heatmap_patway(GSEA_parsing_data = res_path,
               GeneSetName = GeneSetName,
               rld,
               DEG=0)

# HALLMARK_OXIDATIVE_PHOSPHORYLATION
GeneSetName="HALLMARK_OXIDATIVE_PHOSPHORYLATION"
GSEA_file="../GSEA/my_analysis.Gsea.1461755156354.Hallmark/HALLMARK_OXIDATIVE_PHOSPHORYLATION.xls"
res_path = GSEA_parse(GSEA_file,AllGenes)

# Heatmap
heatmap_patway(GSEA_parsing_data = res_path,
               GeneSetName = GeneSetName,
               rld,
               DEG=0)

# HALLMARK_BILE_ACID_METABOLISM
GeneSetName="HALLMARK_BILE_ACID_METABOLISM"
GSEA_file="../GSEA/my_analysis.Gsea.1461755156354.Hallmark/HALLMARK_BILE_ACID_METABOLISM.xls"
res_path = GSEA_parse(GSEA_file,AllGenes)

# Heatmap
heatmap_patway(GSEA_parsing_data = res_path,
               GeneSetName = GeneSetName,
               rld,
               DEG=0)

# KEGG_FATTY_ACID_METABOLISM
GeneSetName="KEGG_FATTY_ACID_METABOLISM"
GSEA_file="../GSEA/KEGG.Gsea.1461765083398/KEGG_FATTY_ACID_METABOLISM.xls"
res_path = GSEA_parse(GSEA_file,AllGenes)

# Heatmap
heatmap_patway(GSEA_parsing_data = res_path,
               GeneSetName = GeneSetName,
               rld,
               DEG=0)

```


```{r devel}

```




